---
title: "Introduction to easyLSEA"
author: "David Guardamino Ojeda, MD"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Introduction to easyLSEA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 5,
  out.width = "100%"
)
set.seed(42)
```

## Overview

**easyLSEA** provides biology-aware Lipid Set Enrichment Analysis (LSEA)
for untargeted lipidomics data. It combines two complementary enrichment
engines -- the Kolmogorov-Smirnov (KS) test and the fast gene set
enrichment algorithm (fgsea) -- with a dedicated fatty acid chain analysis
module, all wrapped in a single-function interface.

The typical workflow is:

1. **Annotate** lipid names with class and category information
2. **Enrich** lipid classes and categories using KS and/or fgsea
3. **Analyse** fatty acid chain composition patterns
4. **Visualise** and **export** results

## Installation

```{r install, eval = FALSE}
# From GitHub
remotes::install_github("DavidGO464/easyLSEA")

# Optional: fgsea engine (Bioconductor)
BiocManager::install("fgsea")
```

## The example dataset

easyLSEA ships with `lipid_example`, a synthetic dataset of 200 lipid
species simulating a NASH vs Control comparison. PC and PE species are
enriched in NASH; TG species are depleted. This known structure lets us
verify that the analysis recovers the embedded signal.

```{r data}
library(easyLSEA)

data("lipid_example")
head(lipid_example)
```

```{r data-summary}
table(lipid_example$LipidClass)
```

## Step 1 -- Lipid annotation

`annotate_lipids()` assigns each lipid to its class, LIPID MAPS structural
category, and functional category using hierarchical prefix matching
validated against real VLDL lipidomics data.

```{r annotate}
annotated <- annotate_lipids(
  data      = lipid_example,
  lipid_col = "LipidName",
  verbose   = TRUE
)

# Annotation columns added
annotated[1:4, c("LipidName", "LipidClass", "LipidCategory_LMAPS",
                  "LipidCategory_functional")]
```

The internal classifier covers all major lipid classes encountered in
untargeted lipidomics: glycerophospholipids, sphingolipids, glycerolipids,
fatty acyls, sterols, acylcarnitines, oxylipins, and bile acids. An
optional `lipidAnnotator` backend is available for enhanced annotation
(requires separate installation).

## Step 2 -- Lipid Set Enrichment Analysis

`run_lsea()` tests whether lipid sets (defined by class or category) are
coordinately shifted between groups. It tests three biological levels
simultaneously: `LipidClass`, `LipidCategory_LMAPS`, and
`LipidCategory_functional`.

```{r lsea, message = FALSE}
lsea_out <- run_lsea(
  data      = annotated,
  fc_col    = "logFC",
  pval_col  = "P.Value",
  case_lbl  = "NASH",
  ref_lbl   = "Control",
  engine    = "ks",      # "ks", "fgsea", or "both"
  n_perm    = 500L,      # increase to 2000 for publication
  verbose   = FALSE
)

# KS results sorted by p-value
head(lsea_out$ks[order(lsea_out$ks$KS_pval), ])
```

### Interpreting the KS output

| Column | Meaning |
|---|---|
| `Group` | Lipid set (class or category) |
| `N_group` | Number of lipids in the set |
| `DirectionalScore` | Cohen's d analog -- effect size and direction |
| `KS_pval` | KS test p-value (primary inference) |
| `DS_perm_pval` | Permutation p for the DirectionalScore (secondary) |
| `FDR_LSEA` | Benjamini-Hochberg adjusted p-value |
| `Direction` | "up in NASH" or "up in Control" |
| `ContributingLipids_KS` | Lipids driving the distributional shift |

The `DirectionalScore` is a standardised mean difference (Cohen's d
analog). Positive values indicate higher logFC in the case group (NASH);
negative values indicate lower logFC (depleted in NASH).

### Running both engines

When `engine = "both"`, results from KS and fgsea are merged with a
`Convergence` column indicating sets significant in both methods:

```{r both, eval = FALSE}
lsea_both <- run_lsea(
  data      = annotated,
  fc_col    = "logFC",
  engine    = "both",
  case_lbl  = "NASH",
  ref_lbl   = "Control"
)

# Sets significant in both engines
lsea_both$combined[lsea_both$combined$Convergence == "KS+fgsea", ]
```

## Step 3 -- Chain analysis

`parse_lipid_chains()` extracts fatty acid chain information from lipid
names using class-specific parsing strategies:

- **sn-2 chain**: PC, PE (sn-1 is relatively conserved)
- **N-acyl chain**: SM, Cer, HexCer (sphingoid base excluded)
- **Long format**: TG and PI (one row per acyl chain)
- **Single chain**: LPC, LPE, CAR

```{r chains, message = FALSE}
chains_out <- parse_lipid_chains(
  data      = annotated,
  lipid_col = "LipidName",
  class_col = "LipidClass"
)

# Parsing summary
table(chains_out$summary$status)
```

```{r chains-head}
# Chain observations for parsed lipids
head(chains_out$parsed[, c("LipidName", "LipidClass", "chain_type",
                             "analysis_chain_cl", "analysis_chain_cs")])
```

## Step 4 -- One-call wrapper

For most use cases, `easyLSEA()` runs the complete pipeline in a single
call and returns a structured `easyLSEA_result` object:

```{r wrapper, message = FALSE, warning = FALSE}
result <- easyLSEA(
  data       = lipid_example,
  lipid_col  = "LipidName",
  fc_col     = "logFC",
  case_lbl   = "NASH",
  ref_lbl    = "Control",
  engine     = "ks",
  n_perm     = 500L,
  plots      = FALSE,    # set TRUE to generate ggplot objects
  verbose    = FALSE
)

print(result)
```

### Accessing results

```{r access}
# LSEA table
head(result$lsea$ks[order(result$lsea$ks$KS_pval), ])

# Chain parsing summary
table(result$chains$summary$status)

# Metadata
str(result$meta)
```

## Step 5 -- Export

`export_lsea()` saves all results to a timestamped folder:

```{r export, eval = FALSE}
export_lsea(
  result  = result,
  dir     = "results/",
  prefix  = "NASH_vs_Control",
  format  = c("csv", "excel", "pdf")
)
```

This creates:

```
NASH_vs_Control_2026-05-22/
  tables/
    lsea_results_ks.csv
    lsea_combined.csv
    chain_results.csv
    chain_parsed.csv
  plots/
    bubble_ks.pdf
    ...
  results.xlsx
```

Excel export requires `openxlsx`; HTML export requires `rmarkdown` and
`knitr` (both installable from CRAN).

## Modular use

Each step can be run independently:

```{r modular, eval = FALSE}
# Annotation only
annotated <- annotate_lipids(data, lipid_col = "LipidName")

# Enrichment only
lsea_out <- run_lsea(annotated, fc_col = "logFC", engine = "ks")

# Chain analysis only
chains_out <- parse_lipid_chains(annotated)

# Plots only (from existing result)
plots <- plot_lsea(lsea_out, case_lbl = "NASH", ref_lbl = "Control")
chain_plots <- plot_chains(chains_out, case_lbl = "NASH", ref_lbl = "Control")
```

## Session info

```{r session}
sessionInfo()
```
